---
title: "R Notebook "
subtitle: "<h4 >Fadil Mohammed Surur 2023 <h4>"
output: html_notebook
code_download: true
number_sections: true
---


Step 1: Load the packages.
```{r}
install.packages('tidyverse')
install.packages('skimr')
library(tidyverse) #wrangle data
library(dplyr) #clean data
library(lubridate)  #wrangle date attributes
library(skimr) #get summary data
library(ggplot2) #visualize data
library(readr)
```


Step2: Load the data sets and conctenate them to one data set

```{r}
 tripdata1<- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata1.csv")
 tripdata2 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata2.csv")
 tripdata3 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata3.csv")
 tripdata4 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata4.csv")
 tripdata5 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata5.csv")
 tripdata6 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata6.csv")
 tripdata7 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata7.csv")
 tripdata8 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata8.csv")
 tripdata9 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata9.csv")
 trpdata10 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata10.csv")
 tripdata11 <- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata11.csv")
 tripdata12<- read.csv("C:/Users/KIIT/pycourse/couresera_data/my/tripdata12.csv")
 
```


concatenate
```{r}
tripdata<-rbind(tripdata1,tripdata2,tripdata3,tripdata4,tripdata4,tripdata5,tripdata6,tripdata7,tripdata8,tripdata9,tripdata10,tripdata11,tripdata12)

```

Step 3: Examine the data frame.

```{r}
head(tripdata)

```

```{r}
dim(tripdata)

```
```{r}
colnames(tripdata)

```

```{r}
summary(tripdata)
```


Step 4: Check for NA values. 

```{r}
sum(is.na(tripdata)) 
#choose not to remove NA because most NA are related to the start and end station names. 
```
Step 5: Drop columns we don't need: start_lat, start_lng, end_lat, end_lng

```{r}
tripdata <- tripdata %>% select(-c(start_lat, start_lng, end_lat, end_lng))
colnames(tripdata)
```

```{r}
head(tripdata)
```

Step 6a:
Add column "ride_length", which is the length of each ride from ended-at minus started_at, and format as HH:MM:SS.

Step 6b:
Add column “day_of_week,”, and calculate the day of the week that each ride started.

```{r}
tripdata <- tripdata %>%
  mutate(
    started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S"),
    ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S"),
    ride_length = ended_at - started_at,
    day_of_week = weekdays(started_at)
  )


```


```{r}
head(tripdata)
```


now lets convert the ride_length from seconds to minutes


```{r}
tripdata$ride_length <- as.numeric(tripdata$ride_length)
tripdata$ride_length <- as.numeric(tripdata$ride_length/60)
head(tripdata)
```

Step 7: Excluding the problematic data and conducting an analysis on ride_length.

# let us check for data with negative ride length

```{r}
tripdata <- tripdata[tripdata$ride_length>0,]
```

#let us check if there is ride_length  more than 1 day (86400 seconds or 1440 mins).

```{r}
sum(tripdata$ride_length > 1440, na.rm = TRUE)
```
in the above case there is 5452 cases which have more than 1440


#check for extreme outliner
```{r}
max(tripdata$ride_length, na.rm = TRUE)
min(tripdata$ride_length, na.rm = TRUE)
```
#41387.25 mins or 11.50 days  aprox for max ride and 1 second for the least ride length


```{r}
#check for mean and median
mean(tripdata$ride_length, na.rm = TRUE)
median(tripdata$ride_length, na.rm = TRUE)
```
```{r}
summary(tripdata$ride_length)
```

Step 8: Aggregate to analyze the data based on user type: member vs casual


```{r}
aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = mean ,na.rm = TRUE)
aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = median ,na.rm = TRUE)
```
#average ride time by each day for members vs casual
```{r}
tripdata$day_of_week <- ordered(weekdays(as.Date(tripdata$started_at)), levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(tripdata$ride_length ~ tripdata$member_casual + tripdata$day_of_week, FUN = mean, na.rm = TRUE)

```

Step 9: Further analysis into the stations, which shows where NA belong to in step 4.

```{r}
head(count(tripdata, start_station_name, member_casual,  rideable_type, sort= TRUE))
head(count(tripdata, end_station_name, member_casual,  rideable_type, sort= TRUE))
```

Step 10: Analyze ridership data by type and weekday

```{r}
tripdata %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
	
```

Step 11: Visualize the number of rides by rider types AND average duration

```{r}
par(mfrow = c(2, 2))

tripdata %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  na.omit() %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

tripdata %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  na.omit() %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```



Step 12: Visualize Member vs Casual on short ride (less and one hour)


```{r}
one_hour_data <- tripdata %>% filter(ride_length < 60)
one_hour_data$ride_length_by_mins <- cut(one_hour_data$ride_length,breaks = 20)

ggplot(data = one_hour_data) + 
  geom_bar(mapping = aes(x = ride_length_by_mins, fill = member_casual)) +
  labs(title = "One hour ride length") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_discrete(labels=c("3", "6", "9", "12", "15", "18", "21", "24", "27", "30", "33", "36", "39", "42", "45", "48", "51", "54", "57", "60"))
```


Step 13: Visualize day of the week riding choices between member vs causal

```{r}


tripdata_clean <- tripdata[!is.na(tripdata$day_of_week), ]  # Filter out NA values in day_of_week

ggplot(data = tripdata_clean) +
  geom_bar(mapping = aes(x = factor(day_of_week), fill = rideable_type)) +
  facet_wrap(~ member_casual) +
  labs(title = 'Riding Choice During Day of the Week', x = 'Day of the Week') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```

Step 14: Check for peak time for bike usage between member vs casual


```{r}
hour_data <- tripdata
hour_data$start_hour <- as.numeric(format(strptime(tripdata$started_at, "%Y-%m-%d %H:%M:%OS"), '%H'))

hour_data <- hour_data[complete.cases(hour_data$start_hour),]  # Remove NA values

ggplot(data = hour_data) + 
  geom_bar(mapping = aes(x = start_hour, fill = member_casual), stat = 'count') + 
  facet_wrap(~factor(day_of_week)) +
  labs(title = "bike usage by starting hour", x = "starting hour") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))



 
```

Step 15: Save as csv for further analysis and visualization in Tableau
```{r}
#the data frame
write_csv(tripdata, "bikedata.csv")
```


```{r}
#total and average weekly rides by rider type
summary_ride_weekly <- tripdata %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  
  
write_csv(summary_ride_weekly, "summary_ride_weekly.csv")

#total and average weekly rides by rider type
summary_ride_weekly_type <- tripdata %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  
  
write_csv(summary_ride_weekly_type, "summary_ride_weekly_type.csv")
  
#total and avg monthly rides by rider type
summary_month <- tripdata %>% 
  mutate(month = month(started_at, label = TRUE)) %>%  
  group_by(month,member_casual) %>%  
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%    
  arrange(month, member_casual)
write_csv(summary_month, "summary_ride_monthly.csv")

#most popular stations
popular_stations <- tripdata %>%
  mutate(station = start_station_name) %>%
  drop_na(start_station_name) %>%
  group_by(start_station_name, member_casual) %>%
  summarise(number_of_rides=n())

write_csv(popular_stations, "popular_stations.csv")

#total membership types and rideable types
total_riders <- data.frame(table(tripdata$member_casual))
total_types <- data.frame(table(tripdata$rideable_type))

write_csv(total_riders, "total_riders.csv")
write_csv(total_types, "total_types.csv")
```


